---
subtitle: "TMA4268 Statistical Learning V2019"
title: "Compulsory exercise 2: Group 31"
author: "Ingrid Sofie Skjetne and Johannes Voll Kolstø"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  #html_document
  pdf_document
   
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)
```

```{r rpackages,eval=TRUE,echo=FALSE}
# install.packages("knitr") #probably already installed
# install.packages("rmarkdown") #probably already installed
# install.packages("ggplot2") #plotting with ggplot
# install.packages("ggfortify")  
# install.packages("MASS")  
# install.packages("dplyr")
#install.packages("corrplot")
library(knitr)
library(rmarkdown)
library(ggplot2)
library(ggfortify)
library(MASS)
library(dplyr)
library(ISLR)
library(leaps)
library(glmnet)
library(corrplot)


set.seed(1)
```

# Problem 1

## a)


The ridge estimator is the value $\hat{\beta}_{R}$ that minimizes  
\begin{align*}
  \text{RSS} + \lambda \sum_{j=1}^p \beta_j^2 &= (y-X\beta)^\text{T}(y-X\beta) + \lambda \beta^\text{T}\beta \\
  &= (y^\text{T} - \beta^\text{T} X^\text{T})(y-X \beta) + \lambda \beta^\text{T}\beta\\
  &= (y^\text{T} - \beta^\text{T} X^\text{T})y - (y^\text{T} - \beta^\text{T} X^\text{T}) X\beta + \lambda \beta^\text{T}\beta\\
  &= y^\text{T}y -\beta^\text{T}X^\text{T}y-y^\text{T}X\beta + \beta^\text{T}X^\text{T}X\beta +\lambda \beta^\text{T}\beta.\\
\end{align*}

The value above is the $\text{RSS}$ plus a term which penalizes large values of $\beta$, which forces the parameter $\beta$ to shrink. In order to find $\hat{\beta}_{R}$, we set the partial derivative of the above expression with respect to $\beta$ equal to zero.

The partial derivative with respect to $\beta$ is 

\begin{align*}
&\frac{\partial}{\partial \beta} \left(  (y-X\beta)^\text{T}(y-X\beta) + \lambda \beta^\text{T}\beta  \right) \\
=& \frac{\partial}{\partial \beta} \left(  y^\text{T}y -\beta^\text{T}X^\text{T}y-y^\text{T}X\beta + \beta^\text{T}X^\text{T}X\beta +\lambda \beta^\text{T}\beta  \right) \\
=& -2X^\text{T}y+2X^\text{T}X\beta +2\lambda I.
\end{align*}

Setting the partial derivative equal to zero yields
\begin{align*}
-2X^\text{T}y+2X^\text{T}X\hat{\beta}_{R} +2\lambda I \hat{\beta}_{R} &= 0\\
\implies -X^\text{T}y+X^\text{T}X\hat{\beta}_{R} +\lambda I\hat{\beta}_{R} &= 0 \\
\implies X^\text{T}X\hat{\beta}_{R} +\lambda I\hat{\beta}_{R} &= X^\text{T}y \\
\implies \hat{\beta}_{R} &= (X^\text{T}X +\lambda I)^{-1}X^\text{T}y
\end{align*}


## b)


The expected value of $\hat{\beta}_{R}$ as a function of $X$ is 
\begin{align*}
(X^\text{T}X +\lambda I)^{-1}X^\text{T}X\beta.
\end{align*}



The covariance matrix of the ridge estimator as a function of $X$ is 
\begin{align*}
&(X^\text{T}X +\lambda I)^{-1}X^\text{T} \text{Var}(Y) ((X^\text{T}X +\lambda I)^{-1}X^\text{T})^\text{T} \\
=& (X^\text{T}X +\lambda I)^{-1}X^\text{T} \sigma^2 I ((X^\text{T}X +\lambda I)^{-1}X^\text{T})^\text{T} \\
=& \sigma^2 (X^\text{T}X +\lambda I)^{-1}X^\text{T}  X((X^\text{T}X +\lambda I)^{-1})^\text{T} \\
\end{align*}


## c)


i. True.
ii. False?
iii. False.
iv. True. 


## d)

Splitting the `College` dataset into training data and test data.
```{r}
train.ind = sample(1:nrow(College), 0.5*nrow(College))
college.train = College[train.ind, ]
college.test = College[-train.ind, ]
```

Next, we apply forward stepwise selection to choose a model. 

```{r, fig.height=7, fig.width=9}
outstate_forw_fit = regsubsets(Outstate ~ . , nvmax = ncol(College)-1, data=College, method="forward")
outstate_summary = summary(outstate_forw_fit)
```


```{r, fig.height=5, fig.width=6, fig.show="hold"}
plot(outstate_summary$adjr2, type="l", xlab="Number of variables", ylab="Adjusted R squared")
axis(side=1, at= c(1:17))
plot(outstate_summary$bic, type="l", xlab="Number of variables", ylab="BIC")
axis(side=1, at= c(1:17))
plot(outstate_summary$cp, type="l", xlab="Number of variables", ylab="Cp")
axis(side=1, at= c(1:17))
```

Looking at the plots displaying $R^2$, $\text{BIC}$ and $C_p$, we can see that model 7 to 11 perform quite similarily, and model 12 to 17 perform quite similarily. Therefore, it is natural to choose either model 7 or model 11, as these seem to perform well given the number of predictors used.

We choose the model using 7 predictors, as this has a high adjusetd $R^2$, low $\text{BIC}$ and low $C_p$, while still having few predictors. 

We extract the coefficients of model 7.
```{r}
coef(outstate_forw_fit, 7)
```
The model is 
$$
\begin{aligned}
\texttt{Outstate} = &-2858.66 + 2681.26 \cdot \texttt{PrivateYes} +0.95 \cdot \texttt{Room.Board}-0.33 \cdot \texttt{Personal} \\
&+35.95 \cdot \texttt{PhD} + 45.36\cdot\texttt{perc.alumni} +0.23 \cdot \texttt{Expend} +27.98\cdot \texttt{Grad.Rate}.
\end{aligned}
$$

To find the MSE on the test set, we first fit a linear model using the variables in the chosen model.

```{r}
outstate_model_7 = lm(data=College, Outstate ~ Private + Room.Board + Personal + PhD + perc.alumni + Expend + Grad.Rate)
```

Now, out of state tuition for the colleges in the test set is predicted using the model.

```{r}
outstate_test_real = college.test[9]
predicted_outstate = predict(outstate_model_7, college.test[-9])
```

Calculating the test mean squared error.

```{r}
residuals = outstate_test_real - predicted_outstate
mse_ridge = mean(as.matrix(residuals^2))
mse_ridge
```
The mean squared error on the test set is `r mse_ridge`. 

## e)

Using the Lasso method to perform model selection using the `College` dataset. 

First converting the training and test data into matrix format.
```{r}
college_predictors_matrix = data.matrix(college.train[-9])
college_response_matrix = data.matrix(college.train[9])

college_predictors_test = data.matrix(college.test[-9])
college_response_test = data.matrix(college.test[9])
```

Now using the lasso method and plotting the coefficients found vs. $\lambda$. 
```{r, fig.height=7, fig.width=9, fig.show="hold"}
grid = 10^seq(10, -2, length=100)

lasso_college = glmnet(college_predictors_matrix, college_response_matrix, lambda=grid)
plot(lasso_college, xvar = "lambda", label=TRUE, xlim=c(-5, 10))
plot(lasso_college, xvar = "lambda", label=TRUE, ylim=c(0, 60), xlim=c(-5, 10))
```
Performing cross validation to find the best value for $\lambda$. 
```{r, fig.height=7, fig.width=9}
cv_outstate <- cv.glmnet(college_predictors_matrix, college_response_matrix, alpha=1, standardize=TRUE, nfolds=10)
cv_outstate
plot(cv_outstate)
bestlambda = cv_outstate$lambda.min
bestlambda
```

Using the value of $\lambda$ which was found using cross validation and finding the mean squared prediction error when using the chosen model on the test set.

```{r}
lasso_outstate_predictions = predict(lasso_college, s=bestlambda, newx=college_predictors_test)

mse_lasso = mean((lasso_outstate_predictions - college_response_test)^2)
mse_lasso
```
Finding the coefficients of the new model.

```{r}
lasso_coef = predict(lasso_college, type="coefficients", s=bestlambda)[1:18,]
lasso_coef
```
As the above print-out shows, no predictors were removed from the model. The MSE on the test set was `r mse_lasso`. 





