---
subtitle: "TMA4268 Statistical Learning V2019"
title: "Compulsory exercise 2: Group 31"
author: "Ingrid Sofie Skjetne and Johannes Voll Kolstø"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 # html_document
  pdf_document
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)
```

```{r rpackages,eval=TRUE,echo=FALSE}
# install.packages("knitr") #probably already installed
# install.packages("rmarkdown") #probably already installed
# install.packages("ggplot2") #plotting with ggplot
# install.packages("ggfortify")  
# install.packages("MASS")  
# install.packages("dplyr")  
library(knitr)
library(rmarkdown)
library(ggplot2)
library(ggfortify)
library(MASS)
library(dplyr)
library(tree)
library(randomForest)
library(gbm)
library(ISLR)
library(caret)
set.seed(1)
```

# Problem 3

## a)
i. False.
ii. True.
iii. True.
iv. False.

## b)

Loading dataset.

```{r}
train.ind = sample(1:nrow(College), 0.5*nrow(College))
college.train = College[train.ind, ]
college.test = College[-train.ind, ]
```

We will create a random forest model to predict the out-of-state tuition variable `Outstate`. First, we find the `mtry` parameter using cross validation.
```{r}
set.seed(1)
control <- trainControl(method="cv", number=10, search="grid")


tunegrid <- expand.grid(.mtry = (1:17)) 
rf_gridsearch <- train(Outstate ~ ., 
                       data = college.train,
                       method = 'rf',
                       metric = 'RMSE',
                       tuneGrid = tunegrid,
                       trControl=control)
best_mtry = unlist(rf_gridsearch$bestTune)
best_mtry
```


The best value for`mtry` was found to be `best_mtry`. Fitting a random forest model to the training set using the value of `mtry` found using cross validation.

```{r}
rf_college = randomForest(Outstate~., data=college.train, mtry=best_mtry,  ntree=500, importance=TRUE)
rf_college
plot(rf_college)
```

Random forest improves on a single pruned regression tree and on bagging by using many trees that are decorrelated by introducing a random limitation on which splits can be made in each tree. Compared to boosting, a random forest model has few tuning parameters. We need only tune `mtry`, the number of possible parameters the model can use in each split in the tree. In boosting, there are three tuning parameters: the number of trees, the shrinkage parameter, and the number of splits in each tree. 

A disadvantage of random forest is that it is less intuitive and interpretable than a single pruned decision tree. However, compared to boosting, there is no large difference in interpretability. 



## c)

